```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

---
title: "MT_GEMSTONE_Terra_Workshop"
author: "Marco Teixeira"
date: "June 16, 2026"
output:
  html_document:
    df_print: paged
---

# Import packages and setup workspace variables

Load required packages. AnVIL lets you read data tables in Terra.

```{r}
#BiocManager::install("AnVILGCP")
library(AnVILGCP)
library(reticulate) # Iteroperability between R and Python
library(googleCloudStorageR) # Functions to interact with Google Buckets

# Data stuff
install.packages("lme4")
library(dplyr)
library(tidyr)
library(lme4)

# Use ape for phylogenetic analysis
#install.packages("ape")
library(ape)

# import the Python FISS API to interact with Terra
fapi <- import("firecloud.api") 
```

Load environment variables with workspace informaion into R variables so that we can more easily use them in the future.

```{r}
project <- Sys.getenv('WORKSPACE_NAMESPACE')
workspace <- Sys.getenv('WORKSPACE_NAME')
bucket <- Sys.getenv('WORKSPACE_BUCKET')

# Define the Google Cloud Storage bucket so we can use the GCP R library
gcs_global_bucket(bucket)
# Authenticate so we can fetch data from Google Cloud Storage
gcs_auth()
```

# Submit the GEMSTONE C. auris workflow

The standard GEMSTONE _C. auris_ workflow uses CladeTyper to determine _C. auris_ clades for each sample. We will use the Terra GUI and the FISS API to run this workflow in Terra. But first, we must load the Terra entity data tables with sample information, and subset them.

## Load required tables

We will now read the _cauris_sample_ and _resident_ data tables from the workspace. Because we are working from within the workspace containing the data tables, we only need to use the `avtable()` function. However, we could also read tables from other workspaces, such as the DSSC one, using `avworkspace("gemstone-dssc/GEMSTONE%20-%20DSSC")`, followed by `avtable(table_name)`.

```{r}
# Print a list of existing data tables
workspace_tables <- avtables()
print(paste("Sample tables present in this workspace are:", 
            as.character(workspace_tables[,1])))

# Load C. auris isolate table
cauris <- avtable("cauris_sample")
# Load table with resident metadata
resident <- avtable("resident")

# Inspect the cauris_sample table
head(cauris)
```

## Submit the workflow
We will run the GEMSTONE _C. auris_ workflow for the 10 samples in the _cauris_sample_ table. 
The table has 11 columns:

* *cauris_sample_id*: the sample ID;
* *raw_reads1* and *raw_reads2*: FASTQ files with Illumina sequencing data, which will be used as inputs to the workflow;
* *project*: the study from which the isolate originated (e.g., CLEAN or GEMSTONE);
* *site*: the swab site;
* *nh_code*: the nursing home from which the isolate was collected;
* *culture_id*: a unique ID for each isolate. Multiple sequencing versions for the same isolate have the same culture ID, but different sample IDs;
* *organism*: the MDRO type;
* *swab_id*: a unique ID for the swab;
* *record_id*: the record ID for the host from which the isolate was collected. The same person or object sampled at different time points has different record IDs;
* *species*: the species or genus, as determined by the lab. Required for the GEMSTONE _C. auris_ workflow.

Before submitting the workflow, we must import the _FungalQC_ workflow to the workspace and define the inputs. After that is done, we can create a sample set for all samples in the `cauris_sample` table and submit the workflow.

```{r}

# First, we need to create an entity set containing all the samples we want to run the workflow for
cauris_set_id <- "GEMSTONE_Terra_workshop"  # Define your custom set name
sample_ids <- cauris$cauris_sample_id

# This dataframe must have two specific columns: the set ID and the targeted entity ID
cauris_set <- data.frame(
  membership_id = rep(cauris_set_id, length(sample_ids)),
  cauris_sample = sample_ids
)

# Format the table names for Terra's data model ('membership:' and '_id' suffixes)
colnames(cauris_set) <- c(
  paste0("membership:", "cauris_sample_set_id"), 
  "cauris_sample"
)

# Upload to Terra
avtable_import_set(
  .data = cauris_set,
  set = "membership:cauris_sample_set_id",
  member = "cauris_sample",
  origin = "cauris_sample",
  namespace = project,
  name = workspace
)

# Define the name of the workflow to run
workflow_name <- "FungalQC"

# Pull existing inputs into an R environment variable
# Note: you can define the inputs and outputs in the Terra GUI and load them here
wf_config <- avworkflow_configuration_get(
  project,
  workflow_name,
  project,
  workspace
)

# Define inputs 
wf_config$inputs[["theiaeuk_illumina_pe.read1"]] <- "this.raw_reads1"
wf_config$inputs[["theiaeuk_illumina_pe.read2"]] <- "this.raw_reads2"
wf_config$inputs[["theiaeuk_illumina_pe.samplename"]] <- "this.cauris_sample_id"

# Change the configuration settings to point to your new set entity type
wf_config$rootEntityType <- "cauris_sample_set"

# Push changes to Terra
fapi$update_workspace_config(project, workspace, project, 
                             workflow_name, wf_config)

# Submit workflows
if (FALSE) {
  submission <- AnVIL::Terra()$createSubmission(
    workspaceNamespace = project,
    workspaceName = utils::URLencode(workspace),
    methodConfigurationNamespace = project,
    methodConfigurationName = wf_config$name,
    entityType = "cauris_sample_set",
    entityName = cauris_set_id,
    useCallCache = TRUE,
    expression = "this.cauris_samples"
  )

  if (submission$status_code == 201) {
    print("Job submitted.")
  } else {
    print("Submission failed: ")
    print(submission)
  }
}

```

# Determine the clade for each isolate
The column `cladetyper_clade` contains the _C. auris_ clade assignments for each isolate. We will determine how many isolates of which clade were sampled, and generate a phylogenetic tree for Clade I isolates.

```{r}
# Reload the cauris table so it includes the workflow outputs
cauris <- avtable("cauris_sample")

# Count the number of isolates per clade and NH
clade_counts <- cauris %>%
  group_by(nh_code, cladetyper_clade) %>%
  summarise(num_isolates = length(read1_clean), .groups = 'drop') %>%
  pivot_wider(names_from = nh_code, values_from = num_isolates, values_fill = 0)

print("Number of clade types per nursing home:")
print(clade_counts)
```

Now, we will create entity sets containing the isolates from each clade and upload them to Terra.

```{r}
for (clade in unique(cauris$cladetyper_clade)) {
  
  # Subset sample IDs from this clade
  clade_ids <- (cauris %>% filter(cladetyper_clade == clade))$cauris_sample_id
  clade_set <- data.frame(
    membership_id = rep(clade, length(clade_ids)),
    cauris_sample = clade_ids
  )
  
  # Format the table names for Terra's data model ('membership:' and '_id' suffixes)
  colnames(clade_set) <- c(
    paste0("membership:", "cauris_sample_set_id"), 
    "cauris_sample"
  )
  
  # Upload to Terra
  avtable_import_set(
    .data = clade_set,
    set = "membership:cauris_sample_set_id",
    member = "cauris_sample",
    origin = "cauris_sample",
    namespace = project,
    name = workspace
  )

}
```

We will generate a phylogenetic tree for Clade I isolates. For this, we will use the [FungalTree workflow](https://dockstore.org/workflows/github.com/broadinstitute/idmp-fungal-pipelines/FungalTree) through the Terra GUI.

First, import this workflow into the Terra workspace; then, define the inputs, select the Clade I isolates from the `cauris_sample_set` table, and run the workflow.

Let's read the outputs from the workflow and plot the phylogenetic tree. The tree, in Newick format, is in the `IqTree2_ml_tree` column.

```{r, fig.width=8, fig.height=6}
# Load the entity set table with the FungalTree outputs
cauris_set <- avtable("cauris_sample_set")

clade <- "Clade1"
newick_col <- "iqtree2_ml_tree" # Column with Newick tree

# Read Newick file from the Google Bucket
newick <- gcs_get_object(
  (cauris_set %>%
    filter(cauris_sample_set_id == clade)) %>%
    pull(newick_col),
  bucket = bucket
)
# Read the first line (the Newick tree) from the stream
newick  <- readLines(textConnection(rawToChar(newick)))[1]

# Use ape to plot the tree
tree <- read.tree(text=newick)

plot(
  tree,
  edge.width = 1,
  label.offset = 0.01,
  cex = 1,
  no.margin = FALSE
)

title(
  paste(c("C. auris", clade, "tree"), collapse=" ")
)
```

# Resident characteristics associated with _C. auris_ clade 1

We will search for resident characteristics associated with colonization with clade I _C. auris_ versus clade III.

First, we need to merge the `cauris_sample` and `resident` tables, based on the `record_id`. The `resident` table contains resident metadata, which we will need for our analyses.

```{r}
# Read the retrospective C. auris table
cauris_retro <- avtable("cauris_sample_retrospective")

data <- left_join(
  cauris_retro,
  resident,
  suffix = c("_cauris", ""),
  by = c("record_id" = "resident_id")
)

# Remove columns with all NaNs
data <- data[ , colSums(is.na(data)) < nrow(data)]

print(
  paste0(
    c("Merged table has", ncol(data), "columns. There are", sum(is.na(data$record_id)),
      "samples without a record ID, and", sum(is.na(data$cladetyper_clade)), 
      "without a C. auris clade assigned."),
    collapse = " "
  )
)

# Remove samples without clade assignments; keep samples with 
# status == preferred, and remove samples without a unique person ID
data <- data %>% 
  drop_na(cladetyper_clade) %>%
  filter(status == "preferred") %>%
  drop_na(person_id)

print(
  paste0(c("Kept", nrow(data), "C. auris isolates."), collapse=" ")
)

```

Let's use a generalized linear mixed model to predict whether a resident is colonized with _C. auris_ clade I or clade III. We model nursing home residents (`person_id` column) and nursing homes (`nh_code` column) as random intercepts.

```{r}
# Define the dependent variable (cladetyper_clade == "Clade1")
data$is_clade1 <- factor(data$cladetyper_clade == "Clade1")

model <- glmer(
  is_clade1 ~ vSNF_bed + wounds + indwelling_device + day_of_stay + year + (1 | person_id) + (1 | nh_code), 
  data = data, 
  family = binomial(link = "logit"),
  control = glmerControl(autoscale=TRUE)
)

summary(model)
```